---
title: "Biostat 213 Homework 2"
author: "Zixi Zhang"
date: "2022-10-15"
output:
  pdf_document: default
  html_document: default
---
<!-- Setup -->
<style type="text/css">
body{ /* Normal  */
      font-size: 17px;
      font-family: "Times New Roman";
  }
h1,h2,h3,h4,h5,h6{
  font-family: "Times New Roman";
}
</style>
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(ggplot2)
library(tidyverse)
library(gridExtra)
library(grid)
```


**Assignment**

Use the accept-reject algorithm to simulate a Gamma distribution



Make sure you set your seed so you can reproduce your results.

**Gamma Distribution**

Probability Density Function:

for $\alpha, \beta >0$ and $x > 0$:

$$
f(x) = \frac{\beta^\alpha x^{\alpha-1} \exp(-\beta x)}{\Gamma(\alpha)}
$$

where $\Gamma(\alpha)$ is the gamma function:

$$
\Gamma(x) = \begin{cases}
  (x-1)!, \text{ if } x \text{ is a positive integer}\\
  \int_0^\infty s^{x-1}\exp(-s) \ ds, \text{ for } x>0
\end{cases}
$$



## Accept-Reject Method to Simulate Gamma Random Variables
Harder:
Since it is fairly simple to simulate a Gamma($\alpha, \beta$) density when $\alpha$ is an integer. It would be useful to use this with the accept-reject method to simulate a more general Gamma distribution; for example, when $\alpha$ is not an integer. 

#Hint: To do this, we need to figure out if a ratio of two Gamma probability distributions have an upper bound.

```{r}
rgamma.ar <- function(n, shape, scale = 1) {
  s <- shape
  s.int <- floor(s)
  b <- s / s.int
  M <-
    gamma(s.int) / gamma(s) * b ** s.int * (s * exp(-1)) ** (s - s.int)
  f <- function(y)
    dgamma(y, shape = s)
  Mg <- function(y)
    M * dgamma(y, shape = s.int, rate = 1 / b)
  acpt <- 0
  total <- 0
  X <- numeric(n)
  while (acpt < n) {
    total <- total + 1
    Y <- sum(-b * log(runif(s.int)))
    if (runif(1) <= f(Y) / Mg(Y)) {
      acpt <- acpt + 1
      X[acpt] <- Y
    }
  }
  return(list(
    X = scale * X,
    theoretical_rate = 1 / M,
    observed_rate = acpt / total
  ))
}
```


```{r}
#set shape = 5.5, rate = 1
set.seed(2016)
o <- rgamma.ar(1000, shape = 5.5)
#get acceptance probability
o[-1]
```


```{r}
# create list to hold our two plots
plts <- list()

# construct histogram of generated values
plts[[1]] <- tibble(sim = o$X) %>%
  ggplot() + geom_density(aes(x = sim)) + labs(x = "x", y = "Frequency", title =
                                                 "Generated Gamma density") +
  xlim(0, 18) + ylim( c(0, 0.2)) + theme_bw()

# get the exact values of the PDF
gam <- dgamma(seq(0, 18, 0.1), 5.5, rate = 1)
# construct line chart of exact pdf
plts[[2]] <- tibble(x = seq(0, 18, 0.1), pdf = gam) %>%
  ggplot() + geom_line(aes(x = x, y = pdf)) + labs(x = "x", y = "f(x)", title =
                                                     "Exact Gamma density") + 
  xlim(0, 18) + ylim(c(0, 0.2)) + theme_bw()

#arrange two plots
grid.arrange(grobs = plts, nrow = 1)
```



<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
